Software Citation Principles

### 2016-04-05 sc-principles/software-citation-principles.tex
Section 4 suggested changes.

With close to 50 working group participants (see Appendix~\ref{app:wg_members}) representing a range of research domains, the working group was tasked to
document existing practices in their respective communities. A total of 47 documents were submitted by working group
participants, with the life sciences, astrophysics, and geosciences being particularly well-represented in the submitted
resources.

@@

While not usually expressed as software citation principles, a number of groups have developed community guidelines around software and data citation. Van de Sompel et al.~\cite{VandeSompel2004} argue for registration of all units of scholarly communication, including software. In ``Publish or be damned? An alternative impact manifesto for research software''~\cite{ssi-publish-or-be-damned}, Chue Hong lists nine principles as part of `The Research Software Impact Manifesto'.

@@

Most notable of the community efforts are those of WSSSPE~\cite{wssspe} and
SSI~\cite{ssi-workshops}, who between them have run a series of workshops aimed
at gathering together community members with an interest in 1) defining the set of problems
related to the role of software and associated people in research settings, particularly academia,
2) discussing potential solutions to those problems, and 3) beginning to work on implementing
some of those solutions.
%\asnote{Not sure this is a sufficient one-line
%summary but couldn't think of anything better.} \katznote{I've edited it a bit}
In each of the three years that
WSSSPE workshops have run thus far, the participants have produced a report~\cite{WSSSPE1,WSSSPE2,WSSSPE3}
documenting the topics covered. Section 5.8 and Appendix J in the WSSSPE3 report~\cite{WSSSPE3}
has some preliminary work and discussion particularly relevant to this working
group.
Up
In addition, a number of academic publishers such as APA~\cite{APA-guidelines} have recommendations for submitting authors on how to cite software and journals such as F1000Research~\cite{F1000}, SoftwareX~\cite{softwareX}, and Open Research Computation~\cite{softwareX} allow for submissions entirely focussed on research software.

@@

One approach to increasing the ``citability'' of software is to encourage the
submission of papers in standard journals describing a piece of research
software, often known as software papers (see \S\ref{sec:software_papers}).
While some journals (e.g., Transactions on Mathematical Software [TOMS], Bioinformatics, Computer Physics Communications, F1000Research) have traditionally accepted
software submissions, the American Astronomical Society (AAS) have recently
announced they will accept software papers in their journals~\cite{aas-sofware-papers}.
Professional societies are in a good position to change their respective communities, as the publishers of journals and conveners of domain-specific conferences; as publishers they can change editorial policies (as AAS has done) and conferences are an opportunity to communicate and discuss these changes with their communities.

In astronomy and astrophysics: The Astronomy Source Code Library
(ASCL)~\cite{ascl}, is a website dedicated to the curation and indexing of
software used in the astronomy-based literature. In 2015, the AAS and GitHub
co-hosted a workshop~\cite{aas-software-index} dedicated to software citation,
indexing and discoverability and more recently, a Birds of a Feather session
was held at the Astronomical Data Analysis Software and Systems (ADASS) XXV
conference~\cite{2015arXiv151207919A} which included discussion of software
citation.

In the life sciences: In May 2014, the NIH held a workshop aimed at helping the biomedical community discover, cite and reuse software written by their peers. The primary outcome of this workshop was the Software Discovery Index Meeting Report~\cite{software-discovery-index}. The authors of the report cite discuss what framework would be required for supporting a Software Discovery Index including the need for unique identifiers, how citations to these would be handled by publishers and the critical need for metadata to describe software packages.

In the geosciences: The Ontosoft~\cite{ontosoft} project describes itself as `A Community Software Commons for the Geosciences'. Lots of attention given to the metadata
required to describe, discover and execute research software. NSF geosciences workshop around data lifecycle, management and citation~\cite{nsf-geo-data}. This report includes many recommendations for data citation.
\LJHnote{ The Seismological Research Letters, Electronic Seismologists publishes papers on software.} %text needs to be cleaned up LJH%

@@

Producing detailed specifications and recommendations for possible metadata standards to support software citation was not within the scope of this working group. However some discussion on the topic did occur and there is significant interest in the wider community to produce standards for describing research software.

Notable examples around metadata include DOAP~\cite{DOAP}, an early metadata term set used by the Open Source Community, as well as more recent community efforts like the AAS workshop~\cite{aas-software-index}, Project CRediT~\cite{casrai-credit}, Ontosoft~\cite{ontosoft}, CodeMeta~\cite{codemeta}, or the terms and classes defined at \url{schema.org} related to the \url{https://schema.org/SoftwareApplication} class.

>>>>

With approximately 50 working group participants (see Appendix~\ref{app:wg_members})
representing a range of research domains, the working group was tasked to
document existing practices in their respective communities. A total of 47
documents were submitted by working group participants, with the life sciences,
astrophysics, and geosciences being particularly well-represented in the
submitted resources.

@@

While not usually expressed as software citation principles, a number of groups
have developed community guidelines around software and data citation. Van de
Sompel et al.~\cite{VandeSompel2004} argue for registration of all units of
scholarly communication, including software. In ``Publish or be damned? An
alternative impact manifesto for research
software''~\cite{ssi-publish-or-be-damned}, Chue Hong lists nine principles as
part of ``The Research Software Impact Manifesto.''

@@

Most notable of the community efforts are those of WSSSPE~\cite{wssspe} and
SSI~\cite{ssi-workshops}, who between them have run a series of workshops aimed
at gathering together community members with an interest in
\begin{enumerate*}[series=InlineList, before=\hspace{-0.6ex}]
    \item defining the set
of problems related to the role of software and associated people in research
settings, particularly academia,
    \item discussing potential solutions to those
problems, 
\item beginning to work on implementing some of those solutions.
  \end{enumerate*}
%\asnote{Not sure this is a sufficient one-line
%summary but couldn't think of anything better.} \katznote{I've edited it a bit}
In each of the three years that WSSSPE workshops have run thus far, the
participants have produced a report~\cite{WSSSPE1,WSSSPE2,WSSSPE3} documenting
the topics covered. Section~5.8 and Appendix~J in the WSSSPE3
report~\cite{WSSSPE3} has some preliminary work and discussion particularly
relevant to this working group.  In addition, a number of academic publishers
such as APA~\cite{APA-guidelines} have recommendations for submitting authors on
how to cite software and journals such as F1000Research~\cite{F1000},
SoftwareX~\cite{softwareX}, and Open Research Computation~\cite{softwareX} allow
for submissions entirely focussed on research software.

@@

One approach to increasing software ``citability''   is to encourage the
submission of papers in standard journals describing a piece of research
software, often known as software papers (see \S\ref{sec:software_papers}).
While some journals (e.g., Transactions on Mathematical Software (TOMS),
Bioinformatics, Computer Physics Communications, F1000Research) have
traditionally accepted software submissions, the American Astronomical Society
(AAS) has recently announced they will accept software papers in their
journals~\cite{aas-sofware-papers}. Professional societies are in a good
position to change their respective communities, as the publishers of journals
and conveners of domain-specific conferences; as publishers they can change
editorial policies (as AAS has done) and conferences are an opportunity to
communicate and discuss these changes with their communities.

In astronomy and astrophysics: The Astronomy Source Code Library
(ASCL)~\cite{ascl}, is a website dedicated to the curation and indexing of
software used in the astronomy-based literature. In 2015, the AAS and GitHub
co-hosted a workshop~\cite{aas-software-index} dedicated to software citation,
indexing and discoverability in astrophysics. More recently, a Birds of a Feather session was
held at the Astronomical Data Analysis Software and Systems (ADASS) XXV
conference~\cite{2015arXiv151207919A} which included discussion of software
citation.

In the life sciences: In May 2014, the NIH held a workshop aimed at helping the
biomedical community discover, cite, and reuse software written by their peers.
The primary outcome of this workshop was the Software Discovery Index Meeting
Report~\cite{software-discovery-index}. The authors of the report discuss
what framework would be required for supporting a Software Discovery Index
including the need for unique identifiers, how citations to these would be
handled by publishers, and the critical need for metadata to describe software
packages.

In the geosciences: The Ontosoft~\cite{ontosoft} project describes itself as ``A
Community Software Commons for the Geosciences.'' Much attention was given to the
metadata required to describe, discover, and execute research software. NSF sponsored
Geo-Data Workshop 2011~\cite{nsf-geo-data} revolved around data lifecycle, management, and
citation. The workshop report includes many recommendations for data
citation. \LJHnote{ The Seismological Research Letters, Electronic Seismologists
publishes papers on software.} %text needs to be cleaned up LJH% dy

@@

Producing detailed specifications and recommendations for possible metadata
standards to support software citation was not within the scope of this working
group. However some discussion on the topic did occur and there was significant
interest in the wider community to produce standards for describing research
software metadata.

Notable examples around metadata include DOAP~\cite{DOAP}, an early metadata
term set used by the Open Source Community, as well as more recent community
efforts like the AAS workshop~\cite{aas-software-index}, Project
CRediT~\cite{casrai-credit}, Ontosoft~\cite{ontosoft}, CodeMeta~\cite{codemeta},
or the terms and classes defined at \href{https://schema.org}{www.schema.org} related to the
\href{https://schema.org/SoftwareApplication}{\texttt{SoftwareApplication}} class.
